\chapter*{Publications}

\textbf{Publications related to the current thesis:}
\section*{Journal Papers}

\begin{itemize}
    \item A. Pister, P. Buono, J.-D. Fekete, C. Plaisant, and P. Valdivia, “Integrating Prior Knowledge in Mixed-Initiative Social Network Clustering,” IEEE Transactions on Visualization and Computer Graphics, vol. 27, no. 2, pp. 1775–1785, Feb. 2021, doi: 10.1109/TVCG.2020.3030347.
    \item A. Pister, C. Prieur, J.-D. Fekete, ``ComBiNet: Visual Query and Comparison of Bipartite Multivariate Dynamic Social Networks.'' Computer Graphics Forum. Submitted in fast track.
\end{itemize}

\section*{Workshop Paper}

\begin{itemize}
    \item A. Pister, N. Dufournaud, P. Cristofoli, C. Prieur, and J.-D. Fekete, “From Historical Documents To Social Network Visualization: Potential Pitfalls and Network Modeling,” presented at the IEEE Visualization Conference (VIS) 2022, 2022.
\end{itemize}


\section*{Poster}

\begin{itemize}
    \item  A. Pister, C. Prieur, and J.-D. Fekete, “Visual Queries on Bipartite Multivariate Dynamic Social Networks,” 2022. doi: 10.2312/evp.20221115.
\end{itemize}


\vspace{5em}

%Publications not related to the current thesis:
%\section*{Publications not related to the current thesis:}
\textbf{Publications not related to the current thesis:}
\section*{Journal Papers}
\begin{itemize}
    \item N. Tovanich, A. Pister, G. Richer, P. Valdivia, C. Prieur, J.-D. Fekete, P. Isenberg, “GraphletMatchMaker: Visual Analytics Approaches to Graph Matching in Cybersecurity Communities.” IEEE Visual Analytics Science and Technology, VAST Challenge, Oct. 2020. Available: https://hal.archives-ouvertes.fr/hal-02986447
    \item S. D. Bartolomeo, A. Pister, P. Buono, C. Plaisant, C. Dunne, and J.-D. Fekete, “Six Methods for Transforming Layered Hypergraphs to Apply Layered Graph Layout Algorithms,” Computer Graphics Forum, vol. 41, no. 3, p. 1467, Jun. 2022, doi: 10.1111/cgf.14538.
    \item G. Richer, A. Pister, M. Abdelaal, M. Sedlmair, J.-D. Fekete, D. Weiskopf, ``Scalability in Visualization.'' IEEE Transactions on Visualization and Computer Graphics. Submitted.
\end{itemize}
